\thetitle

### 2008-06-29 paper.tex
Final tweaking: now down to 18 pages.

Fundamentally, uniform, marginal and conditional models are applied so that network traffic behavior can be represented using a manageable collection of empirical or analytical statistical distributions.
Reduction in representation size and complexity makes analysis and generation feasible.
From this perspective, three questions immediately present themselves:
\begin{enumerate}
\item Do these assumptions distort the metrics we are trying to analyze, replicate, and predict?
\item Can we represent ``unreduced'' network behavior such that analysis and generation remain tractable?
\item Are there better ways to simplify and reduce the representation of network behavior?
\end{enumerate}
This paper answers all three questions. We begin, however by answering the second problem and  
using our solution to address the other two questions.

@@

\newfootnote{\pbdnote}{There are many schemes to correct the discrepancy in bytes when $\pkts$ does not divide $\mathsym{bytes}$ evenly. The simplest is to ignore it; we leave more complex schemes to the reader.}

@@

\subsubsection{Uniformity}

@@

\subsubsection{Marginality}

@@

\subsubsection{Conditionality}

@@

The remainder of this section proceeds as follows. In \Section{traffic-models} we present the collection of traffic models we will evaluate. In \Section{trace-data} we describe the trace data that serves as the control for our experiments. Our simulation methodology is detailed in \Section{simulations}, and the performance metrics used for statistical comparison of control and test experiments are described in \Section{performance-metrics}.
% Finally, \Section{measures-of-error} is an in-depth mathematical discussion of the appropriate measure of error for the pairs of metric values that are produced by the paired differential simulation methodology.

@@

Another potential concern about the approach of linear representation is the additional computational overhead it may induce in experimental settings.
Large matrix representations of traffic models are certainly more computationally costly than simpler models.
Currently, however, realistic local-area traffic modeling is an unsolved problem.
The purpose of this work is to establish how traffic can be represented and generated realistically at all, not how it can be done most efficiently.
We observe, however, that generating application traffic for a simulation takes orders of magnitude less \caps{CPU}-time than running the simulation itself.
Therefore, traffic model generation complexity is considered a non-issue.
A related issue is the significant simulation time incurred when experiments use realistic volumes of traffic.
Indeed, this is a problem:
we have run our simulations on a 64-node Itanium cluster, and the experiments still take weeks to complete.
However, with the current state of the art in simulators and \caps{CPU} power, this problem is unavoidable.
Including model preparation times, running simulations with uniform or marginal models takes negligibly less time than running simulations using realistic models like the \caps{GMM}.
One of the possible effects of this research, however, is the potential ability to classify and catalog an realistic set of ``exemplary'' traffic scenarios for various applications.
Such a catalog is far from existing, but its existence would allow researchers to run far fewer simulations to achieve infinitely more reliable results.

@@

It remains to show how we can simplify and reduce the representation of network behaviors without making unwarranted and manifestly false assumptions of uniformity of marginality of network behaviors.
In \Section{modeling-concepts}, we observed that application of uniformity and marginality assumptions to traffic instances is equivalent, in the \caps{GMM}, to right and left multiplication, respectively, by matrices of rank one. These transformations destroy almost all of the information in the original traffic matrix.
Conditional models are equivalent to left-multiplication by somewhat less degenerate matrices, but they achieve realism only by approaching the \caps{GMM} itself.
In this section we describe matrix factorization, which provides an approximate low-rank representation of the original traffic.
This technique not only allows us to model and analyze traffic without destroying the information content, but also leads us to a deep analytical understanding of network behaviors that matches our intuitive understanding of typical modes of network behavior.

We present here preliminary results from applying nonnegative matrix factorization (\caps{NMF}) to the packet behaviors of real flows. Recall from \Section{general-matrix-model} that the matrices $\mat{Z}$ and $\mat{V}$ represent the packet size and inter-packet interval histograms of a collection of flows, with each row corresponding to a flow, and each column corresponding to a packet size or a range of inter-packet intervals. We apply \caps{NMF} to the concatenation of these matrices, $\mat{P}=[\mat{Z}~\mat{V}] \in \R^{f \times (d_z+d_v)}$. \caps{NMF} attempts to approximate $\mat{P}$ as the product of nonnegative matrices:
\begin{align}
\mat{P} \approx \mat{W}\mat{\Sigma}\mat{H},
\end{align}
where $\mat{W} \in \R^{f \times b}$ and $\mat{H} \in \R^{b \times (d_z+d_v)}$ are both row-stochastic, while
$\mat{\Sigma} \in \R^{b \times b}$ is diagonal with descending diagonal entries.
$b\in\N$ is the rank of this factorization, which must be determined somehow as well.
What is the interpretation of such a factorization?
The rows of $\mat{H}$ are \emph{basic behaviors}---pairs of packet size and inter-packet interval distributions, such that each flow's packet behavior can be approximated as a linear combination of these basic behaviors.
Specifically, the matrix $\mat{M}\mat{\Sigma}$ gives the mixing coefficients for each flow.
The diagonal entries of $\mat{\Sigma}$ have a special meaning as well:
they are the number of packets associated with each basic behavior.
Thus the basic behaviors are sorted in descending order of how much traffic they explain.
% Note that the factorization does not associate basic behaviors with individual packets;
% rather, total flow behavior is explained as a mixture of basic behaviors, so the fraction of packets in each flow explained by a given behavior is known, while the specific packets are unknown.

The striking results of applying nonnegative matrix factorization to $\mat{P}$ are shown in \Figure{basic-behaviors}. Twelve basic behaviors suffice to explain all the traffic in a ``toy''  sample of 4138 flows. The pairs of packet size and inter-packet interval distributions are almost immediately recognizable as resembling intuitive modes of network behavior. The first basic behavior, for example has all packets near the \caps{MTU}, while intervals are smoothly distributed around a fraction of a millisecond. This is what we would expect from downloading large files; and indeed, the breakdown of packets by port number bears out this interpretation---almost all of the traffic is \caps{HTTP} server traffic, with small slivers of \caps{IMAP} and \caps{SSH} server traffic. The second basic behavior is equally intuitive: tiny packets with frequency distributed around 0.1 seconds---this is all \caps{SSH} traffic, as indicated by the associate pie chart.
The other behaviors are listed in \Table{basic-behaviors} with interpretations.
% This is precisely the behavior we expect from interacting with a remote terminal via keystrokes and screen updates.
The true significance of this result lies in the fact that \caps{NMF} succeeds in classifying flows by network behavior completely automatically, not only without using port numbers, but also without \emph{any} information about the number or kind of plausible interpretations.´
The algorithm not only agrees with our intuition about network behavior, but exceeds it by discovering real but unknown behaviors awaiting interpretation.

>>>>

Fundamentally, uniform, marginal and conditional models are applied so that network traffic behavior can be represented using a manageable collection of empirical or analytical statistical distributions.
Reduction in representation size and complexity makes analysis and generation feasible.
From this perspective, three questions immediately present themselves:
\begin{enumerate}
\item Do these assumptions distort the metrics we are trying to analyze, replicate, and predict?
\item Can we represent ``unreduced'' network behavior so that analysis and generation remain tractable?
\item Are there better ways to simplify and reduce the representation of network behavior?
\end{enumerate}
This paper answers all three questions. We begin, however by answering the second problem and  
using our solution to address the other two questions.

@@

\newfootnote{\pbdnote}{There are many schemes to correct the discrepancy in bytes when $p$ does not divide $b$ evenly. The simplest is to ignore it; we leave more complex schemes to the reader.}

@@

\subsubsection{Uniform Modeling}

@@

\subsubsection{Marginal Modeling}

@@

\subsubsection{Conditional Modeling}

@@

%The remainder of this section proceeds as follows. In \Section{traffic-models} we present the collection of traffic models we will evaluate. In \Section{trace-data} we describe the trace data that serves as the control for our experiments. Our simulation methodology is detailed in \Section{simulations}, and the performance metrics used for statistical comparison of control and test experiments are described in \Section{performance-metrics}.
% Finally, \Section{measures-of-error} is an in-depth mathematical discussion of the appropriate measure of error for the pairs of metric values that are produced by the paired differential simulation methodology.

@@

Another potential concern about the approach of linear representation is the additional computational overhead it may induce in experimental settings.
Large matrix representations of traffic models are certainly more computationally costly than simpler models.
Currently, however, realistic local-area traffic modeling is an unsolved problem.
The purpose of this work is to establish how traffic can be represented and generated realistically at all, not how it can be done most efficiently.
We observe, however, that generating application traffic for a simulation takes orders of magnitude less \caps{CPU}-time than running the simulation itself.
Therefore, traffic generation complexity is considered a non-issue.

A related issue is the significant simulation time incurred when experiments use realistic volumes of traffic.
Indeed, this is a problem:
we have run our simulations on a 64-node Itanium cluster, and the experiments still take weeks to complete.
Given the current state of the art in simulators and \caps{CPU} power, however, this problem is unavoidable.
Including model preparation times, running simulations with uniform or marginal models takes negligibly less time than running simulations using realistic models like the \caps{GMM}.
One of the possible effects of this research, however, may be the ability to classify and catalog an realistic set of ``exemplary'' traffic scenarios for various applications.
Such a catalog is far from existing, but its existence would allow researchers to run far fewer simulations to achieve incomparably more general and reliable results.

@@

It remains to show how we can simplify and reduce the representation of network behaviors without making unwarranted and manifestly false assumptions of uniformity or marginality of network behaviors.
In \Section{modeling-concepts}, we observed that application of uniformity and marginality assumptions to traffic instances is equivalent, in the \caps{GMM}, to right and left multiplication, respectively, by matrices of rank one. These transformations destroy almost all of the information in the original traffic matrix.
Conditional models are equivalent to left-multiplication by somewhat less degenerate matrices, but they achieve realism only by approaching the \caps{GMM} itself.
In this section we describe matrix factorization, which provides an approximate low-rank representation of the original traffic.
This technique not only allows us to model and analyze traffic without destroying the information content, but also leads us to a deep analytical understanding of network behaviors that matches our intuitive understanding of typical modes of network behavior.

We present here preliminary results from applying nonnegative matrix factorization (\caps{NMF}) to the packet behaviors of real flows. Recall from \Section{general-matrix-model} that the matrices $\mat{Z}$ and $\mat{V}$ represent the packet size and inter-packet interval histograms of a collection of flows.
We apply \caps{NMF} to the concatenation of these matrices, $\mat{P}=[\mat{Z}~\mat{V}] \in \R^{f \times (d_z+d_v)}$. \caps{NMF} attempts to approximate $\mat{P}$ as the product of nonnegative matrices:
\begin{align}
\mat{P} \approx \mat{W}\mat{\Sigma}\mat{H},
\end{align}
where $\mat{W} \in \R^{f \times b}$ and $\mat{H} \in \R^{b \times (d_z+d_v)}$ are both row-stochastic, while
$\mat{\Sigma} \in \R^{b \times b}$ is diagonal with descending diagonal entries.
$b\in\N$ is the rank of this factorization, which must be determined somehow as well.
What is the interpretation of such a factorization?
The rows of $\mat{H}$ are \emph{basic behaviors}---pairs of packet size and inter-packet interval distributions, such that each flow's packet behavior can be approximated as a linear combination of these basic behaviors.
Specifically, the matrix $\mat{M}\mat{\Sigma}$ gives the mixing coefficients for each flow.
The diagonal entries of $\mat{\Sigma}$ have a special meaning as well:
they are the number of packets associated with each basic behavior.
Thus the basic behaviors are sorted in descending order of how much traffic they explain.
% Note that the factorization does not associate basic behaviors with individual packets;
% rather, total flow behavior is explained as a mixture of basic behaviors, so the fraction of packets in each flow explained by a given behavior is known, while the specific packets are unknown.

The striking results of applying nonnegative matrix factorization to $\mat{P}$ are shown in \Figure{basic-behaviors}. Twelve basic behaviors suffice to explain all the traffic in a ``toy''  sample of 4138 flows. The pairs of packet size and inter-packet interval distributions are almost immediately recognizable as resembling intuitive modes of network behavior. The first basic behavior, for example has all packets near the \caps{MTU}, while intervals are smoothly distributed around a fraction of a millisecond. This is what we would expect from downloading large files; and indeed, the breakdown of packets by port number bears out this interpretation---almost all of the traffic is \caps{HTTP} server traffic, with small slivers of \caps{IMAP} and \caps{SSH} server traffic. The second basic behavior is equally intuitive: tiny packets with frequency distributed around 0.1 seconds---this is all interactive \caps{SSH} traffic, as indicated by the associate pie chart.
The other behaviors are listed in \Table{basic-behaviors} with interpretations.
% This is precisely the behavior we expect from interacting with a remote terminal via keystrokes and screen updates.
The true significance of this result lies in the fact that \caps{NMF} succeeds in classifying flows by network behavior completely automatically, not only without using port numbers, but also without \emph{any} information about the number or kind of plausible interpretations.´
The algorithm not only agrees with our intuition about network behavior, but exceeds it by discovering real but unknown behaviors awaiting interpretation.
